---
title: "Introduction to the Tidyverse"
subtitle: "Exploring an Opinionated Grammar of R"
author: "Nicholas R. Davis"
institute: "UW-Milwaukee"
date: "7/29/2019"
output:
  ioslides_presentation: 
   logo: "https://raw.githubusercontent.com/tidyverse/tidyverse/master/pkgdown/favicon/apple-touch-icon-120x120.png"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment="", prompt=TRUE, fig.height=5, fig.width=5)
```

<!-- A few notes about this document: 

  You can alter the YAML header items as you see fit. 
  
  If you want to typeset as PDF slides, select that with the knit menu or change the output to
  'beamer_presentation; you may want to include a theme and color from the choices listed here:
  http://deic.uab.es/~iblanes/beamer_gallery/
  
  For example, you could do this in the YAML header:
  
    theme: "Boadilla"
    colortheme: "dove"

  Also, more information on working with Rmd can be found here: 
  https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf

 -->

## What is the 'Tidyverse'?

A set of packages developed together following a common set of principles.

>- "tidy" data philosophy, where each variable has its own column, each observation has its own row
>- code clarity and reproducibility through common functional structure 
>- use of pipe `%>%` to improve code development and readability

<!-- The '>' before the bullets here indicates that I want these to appear on click. -->

## Packages Included

- [ ] `ggplot2`: data visualization
- [x] `dplyr`: data manipulation
- [ ] `tidyr`: modeling and data management
- [ ] `readr`: open and organize the data
- [ ] `purrr`: code optimization and functional programming
- [x] `tibble`: alternative to data.frame class
- [ ] `stringr`: functions for working with string data
- [ ] `forcats`: functions for working with factors
- [x] also, by default includes `magrittr` (source of the pipe operator)

<!-- I show you a way to create checkboxes here; this has been handy in presentations where 
I wrote about multiple literatures or methods but can only talk about a few.  -->

## A Few Words of Caution

The Tidyverse can provide a useful set of tools, but...

>- it is *not* a perfect solution to all our data problems
>- it is *not* always as stable as base-R
>- it is *not* used by (or even liked by) everyone
>- perhaps most importantly, it is *not* a replacement for base-R 

Therefore, do *not* assume...

>- that it is always your best choice for building R-scripts
>- that everyone will inevitably end up being "tidy"
>- that you can avoid learning base-R for general research tasks


## Why Be Tidy-literate?

The Tidyverse provides a powerful set of tools for working with data. 

>- built as a suite of "data science" tools with a focus on importing, manipulating, visualizing data
>- fairly easy to mix tidy and non-tidy code/functions
>- code clarity (and "piping") useful as user-generated functions or data management tasks become more complex

# Setting Up <br /> (Entering the Tidyverse)

## Install the Tidyverse

You can install everything at once (recommended)

<!-- note that I include code here which does not get processed in the R interpreter! -->
```r 
> install.packages("tidyverse")
```
This package is actually many packages wrapped up together for ease of use. 
<!-- By the way, you can make your own package of packages using this:
      https://github.com/mkearney/pkgverse  -->

## Access the Tidyverse

Load the Tidyverse

<!-- this is a code chunk which gets processed by the R interpreter -->
```{r load}
library(tidyverse)
```


<!-- Note that you can also just load which packages you need for a task instead. -->

## Function Masking and dplyr

What about those conflicts?

>- we see that there are two functions in the `dplyr` package which *mask* base-R functions of the same name
>- this means if we want to access the base functions instead of the tidy ones, we need to specify the namespace
>- we could do this with `base::select()`
>- as a general rule, you might want to load the tidyverse *after* all other packages; this will identify the conflicts for you

# magrittr <br /> (Piping hot code)

## The Pipe Operator

The `%>%` operator (from `magrittr`) has a special purpose.

- takes the object/function call result on the left and "passes" it to the right; it does *not* make assignment by itself 
- functions on the right can be passed the left side by adding "." in place of the argument

```{r pipe1}
x <- rnorm(100)
mean(x)
x %>% mean(.)
```

## Pipe Example

```{r pipe2}
# assign Prestige data to object
prestige.data <- carData::Prestige
# use pipe to return brief overview after removing NA values
prestige.data %>% na.omit(.) %>% car::brief(.)
```

# tibble <br /> (Tidy data frames)

## What is a Tibble?

The tidyverse uses tibbles as an alternative to the `data.frame` class. 

>- tibbles, data frames have many similar properties (rectangular data)
>- tibbles are intended to represent the "tidy" data principles by design
>- tibbles respond well to `dplyr` data manipulation methods but coerce easily back to `data.frame` as well

## Load Data as Tibble


```{r tibble}
(prestige.data <- prestige.data %>% as_tibble)
```

## Properties of Tibbles

The Good:

>- automatic "brief" view; just type object name in console
>- can be subsetted using all the familiar operators/indexing methods

The Bad:

>- it is possible to create column classes which are tidy-specific (via `haven`)
>- sometimes older functions cannot directly use tibbles
>- no row names allowed!

## Coercing Tibbles

It is easy to use tibbles with base-R functions which take arguments of class `data.frame`

- this is because tibbles have multiple class attributes

```{r class1}
prestige.data %>% class(.)
```

- where needed, explicit coercing is simple

```{r class2}
prestige.data %>% as.data.frame %>% class
```

## Example

```{r lmex}
prestige.data %>% 
  lm(prestige ~ income + education + women, data=.)
```

# dplyr <br /> (Tidy data management)

## Basic dplyr Functionality

There are many useful functions for working with data in this package.

>- summarize and group cases  
>- manipulate cases and variables  
>- combining and manipulating data sets

## Summarize

Suppose we wanted to find the means of a few variables:

```{r summarize}
prestige.data %>% 
  filter(!is.na(type)) %>%
  summarise_at(vars(education, income, women, prestige), mean)
```

## Summarize by Group

What about means for each level of the factor 'type'?

```{r bygroup}
prestige.data %>% 
  filter(!is.na(type)) %>%
  group_by(type) %>%
  summarise_at(vars(education, income, women, prestige), mean)
```

## Manipulate Variables

Perhaps we want to create a new variable which is a transformation of `education`:

```{r manipulate}
prestige.data %>% 
  mutate(., educ_deviation = 
           (education - mean(education) ) / sd(education) ) %>%
  select_at(., vars(education, educ_deviation) ) %>% 
  summary
```

## Additional Resources

On the web: 

- <https://www.tidyverse.org/>

Books:

- Wickham. H. and G. Grolemund. "R for Data Science." Online: <https://r4ds.had.co.nz/>

Also see:  

- Data management: <https://tinyurl.com/data-transform-sheet>  
- Data import: <https://tinyurl.com/data-import-sheet> 



